A Road Segment Prioritization Approach for Cycling Infrastructure

### 2021-04-10 README.Rmd
paper and figure edits

---
title: Road Segment Prioritization for Bicycle Infrastructure
date: "`r Sys.Date()`"
geometry: margin=2cm
bibliography:
  - mybibfile.bib
  - rlbibfile.bib
#linenumbers: true
numbersections: true
abstract: "Understanding the motivators and deterrents to cycling is essential for creating infrastructure that gets more people to adopt cycling as a mode of transport. This paper demonstrates a new approach to support the prioritisation of cycling infrastructure and cycling network design, accounting for cyclist preferences and the growing emphasis on 'filtered permeability' and 'Low Traffic Neighborhood' interventions internationally. The approach combines distance decay, route calculation, and network analysis methods to examine where future cycling demand is most likely to arise, how such demand could be accommodated within existing street networks, and how to ensure a fair distribution of investment. Although each of these methods has been applied to cycling infrastructure prioritisation in previous research, this is the first time that they have been combined, creating an integrated road segment prioritisation approach. The approach, which can be applied to other cities, as shown in the Appendix, is demonstrated in a case study of Manchester, resulting in cycling networks that balance directness against the need for safe and stress-free routes under different investment scenarios. A key benefit of the approach from a policy perspective is its ability to support egalitarian and cost-effective strategic cycle network planning. \\par\\textbf{Keywords:} cycling networks, low-traffic neighborhoods, routing, transport equity"
# keywords: "cycling, low-traffic neighborhoods, potential demand, routing"
output:
  # bookdown::github_document2:
  bookdown::pdf_document2:
  # # bookdown::word_document2:
      toc: false
      extra_dependencies: ["subcaption", "longtable", "float", "multirow", "placeins", "multicol"]
editor_options: 
  markdown: 
    wrap: 72
---

@@

In this paper, continuity is analysed by looking at the connectivity of
the network through the graph-theoretic concept of *connected
components*. @natera2019data for example,
study the existing cycling network in terms of its disconnected
components. They propose two different algorithms to connect disconnected
components by their most critical links and, in doing so, measure the
size of the growth of the largest connected component as a function of
the kilometers of network added. The concept of connected components is
also at the core of the methodology proposed by @olmos2020data. After
routing the cycling demand onto the network links, they use percolation
theory to filter out the links based on the aggregate flow passing through them, varying the flow threshold for
filtering to identify the minimum flow at which the whole city is
connected by a giant component. While these approaches deal with
continuity better, they look at the network as a whole when attempting
to improve it, and in doing so fail to account for equitable
distribution of infrastructure.

@@

```{=tex}
\begin{align}\label{eq:pcteqn}
     \logit(C_{p}) = & \alpha + \beta_1 d + \beta_2 \sqrt{d} + \beta_3 d^2 + \beta_4 s + \beta_5 ds + \beta_6 \sqrt{d}s 
\end{align}
```

@@

```{r potdemhistograms, fig.cap="Distribution of Potential Cycling Demand", out.width="30%", fig.show='hold', echo=FALSE}
knitr::include_graphics(c(
  "data/Manchester/Plots/histogram_distance_all_vs_cycling.png",
  "data/Manchester/Plots/histogram_distance_all_vs_cycling_potential.png",
  "data/Manchester/Plots/histogram_distance_cycling_potential_vs_current.png"
))
```

```{r desirefacetcycling, fig.cap="Current and Potential Cycling Demand", out.width="60%", fig.pos="H", fig.align="center", echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/desire_facet_cycling.png")
```

@@

\begin{longtable}{lll}
\caption{OSM Road Types}
\label{table:osmroadtypes}\\
\textbf{OSM Road Type } & \textbf{Description}                                                                                                                                            & \textbf{UK Equivalent}                                                                                                   \endfirsthead 
\hline
Motorway                & \begin{tabular}[c]{@{}l@{}}Road open to high-speed \\vehicles only\end{tabular}                                                                                 & Motorway                                                                                                                 \\* 
\hline
Trunk                   & \multirow{5}{*}{\begin{tabular}[c]{@{}l@{}}Roads that don't \\meet motorway criteria, \\in descending order \\of importance and \\through-traffic\end{tabular}} & A-Roads with Primary Status                                                                                              \\* 
\cline{3-3}
Primary                 &                                                                                                                                                                 & A-Roads with Non-Primary Status                                                                                          \\* 
\cline{3-3}
Secondary               &                                                                                                                                                                 & B-roads                                                                                                                  \\* 
\cline{3-3}
Tertiary                &                                                                                                                                                                 & \begin{tabular}[c]{@{}l@{}}Classified unnumbered roads \\OR unclassified busy through roads\end{tabular}                 \\* 
\cline{3-3}
Unclassified            &                                                                                                                                                                 & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Unclassified (Intended for \\local traffic - 60\% of UK roads)\end{tabular}}  \\* 
\cline{1-2}
Residential             & Function is purely residential                                                                                                                                  &                                                                                                                          \\ 
\hline
Service                 & \begin{tabular}[c]{@{}l@{}}Road that provides \\access to a facility\end{tabular}                                                                               &                                                                                                                          \\ 
\hline
Track                   & \begin{tabular}[c]{@{}l@{}}Unpaved - suitable for \\two track vehicles (mostly rural)\end{tabular}                                                              &                                                                                                                          \\ 
\hline
Cycleway                & \begin{tabular}[c]{@{}l@{}}Designated cycleway - \\open to cyclists only\end{tabular}                                                                           &                                                                                                                          \\ 
\hline
Path                    & \begin{tabular}[c]{@{}l@{}}Unpaved - open to \\non-motorized traffic only\end{tabular}                                                                          &                                                                                                                          \\
\hline
\end{longtable}

@@

\begin{longtable}{lrrr}
\caption{Weighting Profiles}
\label{table:weightprofiles}\\
\multirow{2}{*}{\textbf{OSM Road Type}} & \multicolumn{3}{c}{\textbf{Weighting Profile}}                                                 \\*
                                        & \multicolumn{1}{l}{\textit{Unweighted}} & \multicolumn{1}{l}{\textit{Weighted}} & \multicolumn{1}{l}{\textit{Weighted\_2}}  \endfirsthead
Cycleway                                & 1          & 1                                     & 1                                         \\
Path                                    & 1          & 0.9                                   & 0.9                                       \\
Residential                             & 1          & 0.9                                   & 0.9                                       \\
Service                                 & 1          & 0.9                                   & 0.9                                       \\
Tertiary                                & 1          & 0.9                                   & 0.9                                       \\
Track                                   & 1          & 0.9                                   & 0.9                                       \\
Unclassified                            & 1          & 0.9                                   & 0.9                                       \\
Secondary                               & 1          & 0.8                                   & 0.8                                       \\
Primary                                 & 1          & 0.7                                   & \textcolor{red}{0}                        \\
Trunk                                   & 1          & 0.6                                   & \textcolor{red}{0}                        \\
Motorway                                & 1          & \textcolor{red}{0}                    & \textcolor{red}{0}                       
\end{longtable}

\noindent A weighted distance $d_{w}$ for each road
segment is calculated as follows ^[The **dodgr** r package
[@padgham2019dodgr] is used to route cycling demand onto the road
network. The package uses the OpenStreetMaps (OSM) road network and
allows the user to assign weights to roads based on their type. The
routing is done based on weighted shortest paths, with the distance
along each road segment being divided by a factor to obtain the weighted
distance for routing. It is more intuitive to multiply when weighting a
network, but the dodgr package divides by numbers between 0 and 1, which
achieves the same result. For the sake of reproducibility, we stick to
the convention used in the package.]:

@@

```{r flowsfacetunweighted, fig.cap="Flow Results Based on Unweighted Shortest Paths (Manchester)", out.width="90%", fig.show='hold', echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/flows_facet_unweighted_Manchester.png")
```

```{r flowsfacetweighted, fig.cap="Flow Results Based on Weighted Shortest Paths (Manchester)", out.width="90%", fig.show='hold', echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/flows_facet_weighted_Manchester.png")
```

@@

```{r boxplotcircuity, fig.cap="Effect of banning cyclists from trunk and primary roads for all OD pairs (Manchester)", out.width="35%", fig.show='hold', echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/boxplot_weighted_unweighted_distances.png")
```

@@

```{r cyclinghistmanc, fig.cap="Cycling Commuting Distance - Manchester (2011 Census Data)", out.width="30%", fig.show='hold', echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/histogram_distance_cycling.png")
```

@@

```{r communitiesmanchester, fig.cap="Communities based on potential cycling demand between MSOAs (Manchester)", out.width="75%", fig.show='hold', fig.align = "center", echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/communities_alternative_Manchester.png")
```

@@

```{r, eval = FALSE}
1. Identify all links that have segregated cycling infrastructure and add them to the 
   initial solution
2. Identify all links that neighbor links in the current solution
3. Select neighboring link with highest flow and add it to the solution
4. Repeat steps 2 and 3 until all flow is satisfied or investment threshold is met
```

@@

<!-- 1. Identify all links that have segregated cycling infrastructure and add them to the initial solution
2. Identify all links that neighbor links in the current solution
3. Select *from each community* one neighboring link with highest flow and add it to the solution
4. If there are no more neighboring links in a community, select the link with the highest flow in that community, regardless of connectivity, and add it to the solution
5. Repeat steps 2, 3 \& 4 until all flow is satisfied or investment threshold is met -->

@@

knitr::kable(table,
             caption = "Total Person-Km in Different Communities (Manchester)") %>%
  kable_styling(latex_options = "hold_position")

@@

```{r growthtotal, fig.cap="Comparing Overall (Dashed) and Community Level Person-Km Satisfied (Manchester)", out.width="35%", fig.show='hold', fig.align = "center", echo=FALSE}
knitr::include_graphics(c(
  "data/Manchester/Plots/Growth_Results/growth_utilitarian_satisfied_km_both_flow_column.png",
  "data/Manchester/Plots/Growth_Results/growth_egalitarian_satisfied_km_both_flow_column.png"
))
```

@@

```{r growth3MapandBar, fig.cap="Road Segment Priority (left), disaggregated by road type (right) - Egalitarian Growth", out.width="35%", fig.show='hold', fig.align="center", echo=FALSE}
knitr::include_graphics(c(
  "data/Manchester/Plots/Growth_Results/growth_egalitarian_priority_all_FLOW.png",
  "data/Manchester/Plots/Growth_Results/growth_egalitarian_investment_highways_flow.png"
))
```

@@

```{r componentsandGCC, fig.cap="Network Characteristics", out.width="30%", fig.show='hold', fig.pos="H", fig.align = "center", echo=FALSE}
knitr::include_graphics(c(
  "data/Manchester/Plots/Growth_Results/growth_util_egal_components_number_comparisonManchester.png",
  "data/Manchester/Plots/Growth_Results/growth_util_egal_components_gcc_comparisonManchester.png"
))
```

@@

Given that the "most essential activity entailed in the design of
cycle-friendly infrastructure is developing a cycle network"
[@parkin2018designing], we believe that the approach has great
potential to inform investment in cities such as Manchester where there
is political will to invest in cycling long-term. A benefit of the
approach is that it has relatively modest data requirements: only the
road network (from OSM), its topography (from satellite imagery), and
OD data (from the national census in this case) are
needed, opening up the possibility of deploying the approach in other
cities^[The results are easily reproducible for all UK cities, and can
also be reproduced for cities elsewhere given the availability of
commuter data. For instructions on reproducing the results shown in this
paper, see the **README** file in the R folder of the .zip file that
accompanies this paper (to be released as open source software when the
paper is published).].

>>>>

---
title: A Road Segment Prioritization Approach for Cycling Infrastructure
date: "`r Sys.Date()`"
geometry: margin=2cm
bibliography:
  - mybibfile.bib
  - rlbibfile.bib
#linenumbers: true
numbersections: true
abstract: "Understanding the motivators and deterrents to cycling is essential for creating infrastructure that gets more people to adopt cycling as a mode of transport. This paper demonstrates a new approach to support the prioritisation of cycling infrastructure and cycling network design, accounting for cyclist preferences and the growing emphasis on 'filtered permeability' and 'Low Traffic Neighborhood' interventions internationally. The approach combines distance decay, route calculation, and network analysis methods to examine where future cycling demand is most likely to arise, how such demand could be accommodated within existing street networks, and how to ensure a fair distribution of investment. Although each of these methods has been applied to cycling infrastructure prioritisation in previous research, this is the first time that they have been combined, creating an integrated road segment prioritisation approach. The approach, which can be applied to other cities, as shown in the Appendix, is demonstrated in a case study of Manchester, resulting in cycling networks that balance directness against the need for safe and stress-free routes under different investment scenarios. A key benefit of the approach from a policy perspective is its ability to support egalitarian and cost-effective strategic cycle network planning. \\par\\textbf{Keywords:} cycling networks, low-traffic neighborhoods, routing, transport equity"
# keywords: "cycling networks, low-traffic neighborhoods, potential demand, routing"
output:
  # bookdown::github_document2:
  bookdown::pdf_document2:
  # # bookdown::word_document2:
      highlight: null   # to remove code highlighting in r chunks
      toc: false
      extra_dependencies: ["subcaption", "longtable", "float", "multirow", "placeins", "multicol"]
editor_options: 
  markdown: 
    wrap: 72
# elsevier style for referencing (from https://github.com/citation-style-language/styles)
csl: elsevier-harvard.csl
---

@@

In this paper, continuity is analysed by looking at the connectivity of
the network through the graph-theoretic concept of *connected
components*. @natera2019data for example,
study the existing cycling network in terms of its disconnected
components. They propose two different algorithms to connect disconnected components; one connects the two largest components, and the other connects the largest component to its nearest neighbor. By measuring the growth of the largest connected component as a function of the kilometers of network added, they determine that these approaches are more effective at improving cycling network connectivity than random allocation of cycling infrastructure.
The concept of connected components is
also at the core of the methodology proposed by @olmos2020data. After
routing the cycling demand onto the network links, they use percolation
theory to filter out the links based on the aggregate flow passing through them, varying the flow threshold for
filtering to identify the minimum flow at which the whole city is
connected by a giant component. While these approaches deal with
continuity better, they look at the network as a whole when attempting
to improve it, and in doing so fail to account for equitable
distribution of infrastructure.

@@

```{=tex}
\begin{align}\label{eq:pcteqn}
     logit(C_{p}) = & \alpha + \beta_1 d + \beta_2 \sqrt{d} + \beta_3 d^2 + \beta_4 s + \beta_5 ds + \beta_6 \sqrt{d}s 
\end{align}
```

@@

<!-- ```{r potdemhistograms, fig.cap="Distribution of potential cycling demand", out.width="30%", fig.show='hold', echo=FALSE} -->
<!-- knitr::include_graphics(c( -->
<!--   "data/Manchester/Plots/histogram_distance_all_vs_cycling.png", -->
<!--   "data/Manchester/Plots/histogram_distance_all_vs_cycling_potential.png", -->
<!--   "data/Manchester/Plots/histogram_distance_cycling_potential_vs_current.png" -->
<!-- )) -->
<!-- ``` -->

```{r potdemhistograms, fig.cap="Distribution of potential cycling demand", out.width="100%", fig.show='hold', fig.align="center", echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/histogram_distance_cycling_potential_current_all.png")
```

```{r desirefacetcycling, fig.cap="Current and potential cycling demand", out.width="60%", fig.pos="H", fig.align="center", echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/desire_facet_cycling.png")
```

@@

\begin{longtable}{lll}
\caption{OSM road types}
\label{table:osmroadtypes}\\
\textbf{OSM Road Type } & \textbf{Description}                                                                                                                                            & \textbf{UK Equivalent}                                                                                                   \endfirsthead 
\hline
Motorway                & \begin{tabular}[c]{@{}l@{}}Road open to high-speed \\vehicles only\end{tabular}                                                                                 & Motorway                                                                                                                 \\* 
\hline
Trunk                   & \multirow{5}{*}{\begin{tabular}[c]{@{}l@{}}Roads that don't \\meet motorway criteria, \\in descending order \\of importance and \\through-traffic\end{tabular}} & A-Roads with Primary Status                                                                                              \\* 
\cline{3-3}
Primary                 &                                                                                                                                                                 & A-Roads with Non-Primary Status                                                                                          \\* 
\cline{3-3}
Secondary               &                                                                                                                                                                 & B-roads                                                                                                                  \\* 
\cline{3-3}
Tertiary                &                                                                                                                                                                 & \begin{tabular}[c]{@{}l@{}}Classified unnumbered roads \\OR unclassified busy through roads\end{tabular}                 \\* 
\cline{3-3}
Unclassified            &                                                                                                                                                                 & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Unclassified (Intended for \\local traffic - 60\% of UK roads)\end{tabular}}  \\* 
\cline{1-2}
Residential             & Function is purely residential                                                                                                                                  &                                                                                                                          \\ 
\hline
Service                 & \begin{tabular}[c]{@{}l@{}}Road that provides \\access to a facility\end{tabular}                                                                               &                                                                                                                          \\ 
\hline
Track                   & \begin{tabular}[c]{@{}l@{}}Unpaved - suitable for \\two track vehicles (mostly rural)\end{tabular}                                                              &                                                                                                                          \\ 
\hline
Cycleway                & \begin{tabular}[c]{@{}l@{}}Designated cycleway - \\open to cyclists only\end{tabular}                                                                           &                                                                                                                          \\ 
\hline
Path                    & \begin{tabular}[c]{@{}l@{}}Unpaved - open to \\non-motorized traffic only\end{tabular}                                                                          &                                                                                                                          \\
\hline
\end{longtable}

@@

\begin{longtable}{lrrr}
\caption{Weighting profiles}
\label{table:weightprofiles}\\
\multirow{2}{*}{\textbf{OSM Road Type}} & \multicolumn{3}{c}{\textbf{Weighting Profile}}                                                 \\*
                                        & \multicolumn{1}{l}{\textit{Unweighted}} & \multicolumn{1}{l}{\textit{Weighted}} & \multicolumn{1}{l}{\textit{Weighted\_2}}  \endfirsthead
Cycleway                                & 1          & 1                                     & 1                                         \\
Path                                    & 1          & 0.9                                   & 0.9                                       \\
Residential                             & 1          & 0.9                                   & 0.9                                       \\
Service                                 & 1          & 0.9                                   & 0.9                                       \\
Tertiary                                & 1          & 0.9                                   & 0.9                                       \\
Track                                   & 1          & 0.9                                   & 0.9                                       \\
Unclassified                            & 1          & 0.9                                   & 0.9                                       \\
Secondary                               & 1          & 0.8                                   & 0.8                                       \\
Primary                                 & 1          & 0.7                                   & \textcolor{red}{0}                        \\
Trunk                                   & 1          & 0.6                                   & \textcolor{red}{0}                        \\
Motorway                                & 1          & \textcolor{red}{0}                    & \textcolor{red}{0}                       
\end{longtable}

\noindent A weighted distance $d_{w}$ for each road
segment is calculated as follows ^[The **dodgr** r package
[@padgham2019dodgr] is used to route cycling demand onto the road
network. The package uses the OpenStreetMaps (OSM) road network and
allows the user to assign weights to roads based on their type. The
routing is done based on weighted shortest paths, with the distance
along each road segment being divided by a factor to obtain the weighted
distance for routing. It is more intuitive to multiply when weighting a
network, but the dodgr package divides by numbers between 0 and 1, which
achieves the same result (road types with a weight of 0 are not routed on). 
For the sake of reproducibility, we stick to
the convention used in the package.]:

@@

```{r flowsfacetunweighted, fig.cap="Flow results based on unweighted shortest paths (Manchester)", out.width="90%", fig.show='hold', echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/flows_facet_unweighted_Manchester.png")
```

```{r flowsfacetweighted, fig.cap="Flow results based on weighted shortest paths (Manchester)", out.width="90%", fig.show='hold', echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/flows_facet_weighted_Manchester.png")
```

@@

```{r boxplotcircuity, fig.cap="Effect of banning cyclists from trunk and primary roads for all OD pairs (Manchester)", out.width="35%", fig.show='hold', fig.align="center", echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/boxplot_weighted_unweighted_distances.png")
```

@@

```{r cyclinghistmanc, fig.cap="Cycling commuting distance - Manchester (2011 census data)", out.width="30%", fig.show='hold', fig.align = "center", echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/histogram_distance_cycling.png")
```

@@

```{r communitiesmanchester, fig.cap="Communities based on potential cycling demand between MSOAs (Manchester)", out.width="75%", fig.show='hold', fig.align = "center", echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/communities_alternative_Manchester.png")
```
\  
<!-- this prevents text from floating above figure -->

@@

```{r, eval = FALSE, results='asis'}
1. Identify all links that have segregated cycling infrastructure and add them to the 
   initial solution
2. Identify all links that neighbor links in the current solution
3. Select neighboring link with highest flow and add it to the solution
4. Repeat steps 2 and 3 until all flow is satisfied or investment threshold is met
```

@@

<!-- 1. Identify all links that have segregated cycling infrastructure and add them to the initial solution -->
<!-- 2. Identify all links that neighbor links in the current solution -->
<!-- 3. Select *from each community* one neighboring link with highest flow and add it to the solution -->
<!-- 4. If there are no more neighboring links in a community, select the link with the highest flow in that community, regardless of connectivity, and add it to the solution -->
<!-- 5. Repeat steps 2, 3 \& 4 until all flow is satisfied or investment threshold is met  -->

@@

knitr::kable(table,
             caption = "Total person-km in different communities (Manchester)") %>%
  kable_styling(latex_options = "hold_position")

@@

```{r growthtotal, fig.cap="Comparing overall (dashed) and community level person-km satisfied (Manchester)", out.width="35%", fig.show='hold', fig.align = "center", echo=FALSE}
knitr::include_graphics(c(
  "data/Manchester/Plots/Growth_Results/growth_utilitarian_satisfied_km_both_flow_column.png",
  "data/Manchester/Plots/Growth_Results/growth_egalitarian_satisfied_km_both_flow_column.png"
))
```

@@

```{r growth3MapandBar, fig.cap="Road segment priority (left), disaggregated by road type (right) - egalitarian growth", out.width="35%", fig.show='hold', fig.align="center", echo=FALSE}
knitr::include_graphics(c(
  "data/Manchester/Plots/Growth_Results/growth_egalitarian_priority_all_FLOW.png",
  "data/Manchester/Plots/Growth_Results/growth_egalitarian_investment_highways_flow.png"
))
```

@@

```{r componentsandGCC, fig.cap="Network characteristics", out.width="30%", fig.show='hold', fig.pos="H", fig.align = "center", echo=FALSE}
knitr::include_graphics(c(
  "data/Manchester/Plots/Growth_Results/growth_util_egal_components_number_comparisonManchester.png",
  "data/Manchester/Plots/Growth_Results/growth_util_egal_components_gcc_comparisonManchester.png"
))
```

@@

Given that the "most essential activity entailed in the design of
cycle-friendly infrastructure is developing a cycle network"
[@parkin2018designing], we believe that the approach has great
potential to inform investment in cities such as Manchester where there
is political will to invest in cycling long-term. A benefit of the
approach is that it has relatively modest data requirements: only the
road network (from OSM), its topography (from satellite imagery), and
OD data (from the national census in this case) are
needed, opening up the possibility of deploying the approach in other
cities^[The results are easily reproducible for all UK cities, and can
also be reproduced for cities elsewhere given the availability of
commuter data. The appendix shows the results for 3 other UK cities: Birmingham, Leeds and Nottingham. 
For instructions on reproducing the results shown in this
paper, see the **README** file in the R folder of the .zip file that
accompanies this paper (to be released as open source software when the
paper is published).].
